Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-22 Ben_Manuscripts/hdphmm/Draft.tex
added more response.
Added new figures, revisions

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the sticky hierarchical Dirichlet process autoregressive hidden Markov model
  (HPD-AR-HMM) to solute center-of-mass trajectories generated from long molecular dynamics (MD)
  simulations in a cross-linked inverted hexagonal phase lyotropic liquid crystal (LLC)
  membrane in order to automatically detect a variety of solute dynamical states. 
  %MRS: starts to mix interpretation in with the machine learning too early?  You don't really 
  explain how the individual states can be explained at this point  
  %which can be further explained in terms of solute-membrane interactions.

  %MRS: changing to lead with WHY we group them.
  We can better understand the mechanisms controlling these dynamic states 
  by grouping the states identified by the HDP-AR-HMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane.

@@

  % BJC4: even if we could calculate diffusion constants, wouldn't there still be value 
  % in understanding chemical detail?
  % BJC4: It might not be a good idea to blame our need for this modeling on non-linear MSDs, 
  % especially since our MSDs appear to be at or nearing a linear regime.
  % BJC4: seems like we should be addressing the problem as short time scale subdiffusive 
  % dynamics that dictates measurable long time scale diffusion. Although curves become linear,
  % the slopes should be a product of the short time scale motion.
  %MRS: OK, did a pass below, reorganizing it a bit.  See what you think. I think it's OK to talk
  %about subdiffusion because it's general problem, we aren't making claims about this one.

  The ability to efficiently design novel membrane materials for
  selective separations would be greatly enhanced by easily
  interpretable methods for connecting microscopic fluctuations with
  macroscopic observables. If one can use molecular dynamics (MD)
  simulations to explain important experimental metrics such as solute
  flux and selectivity in terms of detailed chemically dependent
  solute motion, then we can more easily bridge the disconnect between
  theory and application. In many simple systems, short-time scale
  dynamics and large scale properties can be connected by the simple
  the relationship between the mean square displacement (MSD) of the
  solutes and the macroscopic diffusion constant.  However, for
  systems which exhibit complex short timescale subdiffusive dynamics,
  one cannot reliably compute diffusion constants because MSDs are often
  non-linear on the timescales accessible to molecular
  simulation. Additionally, direct calculations of diffusion constant
  may obscure the molecular mechanisms controlling transport,
  hindering the ability to understand or modulate these mechanisms in new membrane 
  materials. We are therefore in need of a more flexible approach for modeling solute 
  dynamics which makes less assumptions about the solute's short time scale behavior.

@@

  In this work we attempt 
  %MRS: 
  %to minimize human effort 
  uncover similar physical meaning through a data-driven process. 
  We utilize the sticky hierarchical Dirichlet process autoregressive hidden Markov model 
  (HDP-AR-HMM), a nonparameteric Bayesian classification technique that allows the user to 
  automatically detect and infer the parameters of an unknown number of latent vector
  autoregressive (VAR) modes present in solute center-of-mass MD 
  trajectory.~\cite{fox_bayesian_2010} The sticky HDP-AR-HMM greatly extends the 
  basic framework of simpler but more limited finite hidden Markov modeling approaches.

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm, an expectation maximization technique, is a commonly 
  used way to determine the hidden state sequence and the underlying parameters of
  a standard HMM,~\cite{baum_maximization_1970} though it is also possible to achieve the same
  results using a Bayesian framework.~\cite{scott_bayesian_2002,jasra_markov_2005} 
  %MRS: not entirely clear what purpose of above sentence its - it sort of implied you will use Bayesian framework?
  % BJC5: Baum welch is most widely used for pure HMM but is an expectation maximization algorithm
  % There are pure Bayesian ways to do it. Tried to clarify above.
  % MRS5: there's some implication in the above about what you will use; I tried to make it more general. 
  One can also place priors on the transition probabilities and parameters of the hidden 
  states' emission distributions in order to iteratively infer their values and to
  estimate the hidden state sequence. HMMs have been used in a broad range of applications
  including speech recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}.
  In molecular simulations, HMMs have been used to characterize molecular kinetics
  in order to determine the equilibrium distribution of various observables.~\cite{thayer_hidden_2002,singhal_using_2004,noe_probability_2008,noe_projected_2013}

@@

  %MRS: define what ``this'' is - important claim, make it clear what it is you are claiming.
  To our knowledge, applying the HDP-AR-HMM in order to gain kinetic and
  mechanistic insight from MD simulation trajectories has not been done
  before.
  % this is a novel application and analysis of the sticky HDP-AR-HMM.
  Typical applications of the HDP-AR-HMM center around determining the 
  number of states, the state sequence and occasional analysis of the emission 
  distributions of a time series. The addition of temporal correlation to 
  the inference procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  In our approach, we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, and we cluster and analyze the temporally-correlated 
  parameters of the model in order to learn physical mechanisms and describe
  solute kinetics. We can use the model parameters determined either before 
  or after clustering to generate new 
  time series trajectories which enable projections 
  %MRS5: tweaking
  %of MD simulations on
  of the MD-generated time series to arbitrary length time scales.

  %BJC4: enumerate?
  %BJC4: might be able to condense this since some of it repeats above.
  We aim to address two primary scientific questions with this work:
  \begin{enumerate}
  \item First, we want to know whether it is possible to use the HDP-AR-HMM to 
  help us predict macroscopic transport properties. We use the VAR parameters
  and state transition probability matrix of the HDP-AR-HMM in order to 
  generate stochastic trajectory realizations with behavior similar to that
  exhibited by MD. One can project these realizations onto much longer 
  time scales with computational ease extending into time scales where 
  regular diffusive behavior again occurs.
  \item Second, we want to know if the HDP-AR-HMM algorithm can help
  %MRS5: I actually think passive voice works better here, since we want to know if it helps in general.
  %researchers
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior.
  We use the parameters of the states identified by the HDP-AR-HMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.
  \end{enumerate}

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10\% water by weight. Approximately one third of the water molecules 
  occupy the tail region with the rest near the pore center.%MRS: cite previous paper here on this.

  We chose to study a subset of 4 of the fastest moving solutes from our previous
  work: methanol, acetic acid, urea and ethylene glycol.
  In addition to exploring membrane structural space the most, these solutes have a
  relatively diverse set of chemical functionality. For each solute we created a 
  separate system and to each system we added 6 solutes per pore for a total of 24 solutes. This number 
  of solutes per pore provides a balance of a low degree of interaction between 
  solutes and a sufficient amount of data from which to generate statistics on the
  time scales which we simulate. Further details on the setup and equilibration of
  these systems are detailed in our previous work.\cite{coscia_chemically_2019}
  %MRS5: might want to mention above that there was no sign of clustering of solutes (if that is true)

@@

  We use the sticky HDP-AR-HMM in order to assign the time-correlated positional 
  fluctuations of solute center-of-mass MD trajectories to discrete dynamical modes while 
  simultaneously inferring the VAR parameters of those modes. We use the sticky model
  in order to emphasize self-transitions since we expect solutes to stay trapped in
  states for appreciable amounts of time. 
%  We analyze solute motion in terms of their
%  cylindrical coordinates, $(r, z)$, with the $z$ axis oriented along the membrane 
%  pores and the radial dimension, $r$, measured from the closest pore center.

@@

  %\subsubsection*{Parameterizing the Hidden States}\label{method:var_params}
  \textit{Parameterizing the Hidden States}: We describe the dynamics of each 
  state visited by solutes in our MD simulations using a first order vector 
  autoregressive (VAR(1)) model. In general, a VAR($r$) process is characterized by a
  vector of observations in a time series that are linearly dependent on $r$ previous
  values of the time series vector:
  \begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  \label{eqn:var}
  \end{equation}
  Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  process is further characterized by a shift in the mean of each dimension by the
  vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  covariance, $\Sigma$. We limited our analysis to an autoregressive order of $r=1$.
  This means that we only parameterize $A_1$. To simplify notation, we will call it $A$. 
  We used a conjugate matrix-normal inverse-Wishart (MNIW) prior on parameters 
  $A$ and $\Sigma$ and a Gaussian prior on $\mathbf{c}$ in order to efficiently infer their 
  values.~\cite{fox_nonparametric_2009}

%  \subsubsection*{Applying the HDP-AR-HMM to MD Trajectories}

@@

  We ran 2000 iterations of the HDP-AR-HMM procedure on each trajectory which
  was sufficient in order to arrive at converged state sequences and VAR parameter
  estimates. By our criteria, the procedure generally converged within 500--1000 iterations. We 
  evaluated convergence based on when the VAR parameter estimates plateaued. 
  We detected plateaus using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of the individual parameters (Section~\ref{S-section:convergence}
  of the Supporting Information). Convergence of the VAR parameters implies that
  the state sequence is relatively stable since the inferred parameters are 
  sensitive to the inputs to the MNIW conjugate priors. The precise boundaries of 
  the state segments can fluctuate; therefore, in states with a low number of 
  emissions,  the variance in $A$ and $\Sigma$ parameters. 
  %MRS5: grammar in previous sentence unclear?  I tried to clean up, but failed.
  In these cases, plateauing
  of the mean in each dimension tends to be a more reliable indicator of convergence (see 
  Figure~\ref{S-fig:convergence3d} of the Supporting Information).

@@

  There are alternative ways to cluster the data dependent on one's goals. Our goal is
  to derive the simplest set of parameters which adequately describe solute motion. 
  However, one may want a more detailed understand of specific dimensions of the 
  clustering dataset. In this case one can cluster on one or more dimensions independently
  and then enumerate the possible clusters. This would ensure very little overlap of
  clusters but may result in a large number of them. Once one has achieved a better 
  understanding of solute-membrane interactions, perhaps by using the HDP-AR-HMM, it may be
  beneficial to incorporate other physically motivated variables into the clustering
  procedure. 
  %For example, one can gain a thorough understanding of the dynamics of
  %different kinds of trapping behavior related to hydrogen bonding and sodium ion association.
  %MRS: last sentence is a little weird. Audience may be left without knowing how this could be achieved.

@@

  \textit{Obtaining Parameters of Clustered States}: We remapped and fixed the
  state sequence based on the cluster assignments and 
  parameterized the clusters using the inference component of the HDP-AR-HMM algorithm.
  First, we modified the ($x$, $y$, $z$) solute trajectories so that they had a 
  mean of zero, leaving only the fluctuations. We took advantage of the
  rotational invariance of the system by rotating the coordinates about the pore 
  axis such that the mean ($x$, $y$) vector of each separate segment was aligned with the 
  $x$ axis, forcing the $y$ component to zero. Next, we subtracted the means from 
  each separate segment in the $x$ and $z$ dimensions. We used
  the inference component of the HDP-AR-HMM algorithm on this modified trajectory to 
  infer the clustered state parameters by fixing the clustered state sequence.

@@

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}
  %BJC4: reworked this section
  %MRS: should explain why here there are two approaches.
  \textit{Generating Stochastic HDP-AR-HMM Trajectory Realizations}: We used two 
  different methods to stochastically generate trajectories because generating
  trajectories from clustered parameter sets can give results that are qualitatively 
  and quantitatively different from those generated by unclustered parameter sets.
  In our discussion we evaluate both methods:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
    \item For each independent MD trajectory, we generated stochastic trajectory
    realizations using the final parameters of the associated HDP-AR-HMM. This 
    results in 24 separate ensembles of trajectories for each solute studied. 
    \item Using the clustered parameter sets, we generated a single ensemble
	of trajectories. Realizations are representative of the combined behavior
	of all solute trajectories.
  \end{enumerate}

@@

  %BJC4: new section below
  \subsection{Predicting Selectivity}\label{method:selectivity}

@@

  Solute flux, as written, can be expressed in terms of the product of solute 
  diffusivity, $D_i$, within the membrane and a sorption coefficient, $K_i$, 
  that quantifies the ability of the solute to move between the bulk solution
  and and the membrane:
  \begin{equation}
    J_i = \frac{D_i K_i}{l}\Delta C
  \end{equation}
  where $l$ is the membrane thickness. For our purposes, $K_i$ and $l$ are 
  irrelevant since we do not study pore entry effects or flux across finite
  length pores. Therefore, in order to calculate selectivity, with respect
  to quantities we can study with our MD simulations, it is only necessary that
  we compare the ratio of diffusion constants:
  \begin{equation}
    S_{ij} = \frac{D_i}{D_j}
    \label{eqn:selectivity_diffusivity}
  \end{equation}
  It can be shown on time scales where motion is uncorrelated, that the diffusion
  constant is proportional to the slope of the linear region of the MSD 
  curve:~\cite{einstein_investigations_1956}
  \begin{equation}
    \overline{z^2(\tau)} = 2 n D t
    \label{eqn:msd_D}
  \end{equation}
  where $n$ is the number of spatial dimensions (1 in our case). In general, the 
  MSD of the realizations of our models enters a linear regime near a 5 ns time lag
  (see Figure~\ref{S-fig:MET_nclusters} of the Supporting Information for example).
  Since MSD sampling decreases with increasing time lag, longer trajectory 
  realizations help to extend the linear region.
  %BJC: need to confirm these details
  For all solutes, we generated 200 ensembles of 24 10 $\mu s$ trajectories and 
  estimated $D$ from the slope fit between 5 ns and 1000 ns. We used weighted 
  %MRS5: is 5 ns to low?
  least squares to fit the slope with weights determined by the number of time 
  lags used to compute the average ($T - \tau$ in equation~\ref{eqn:tamsd}). 
  We bootstrapped error bars for the final reported selectivities. For each 
  bootstrap trial, we sampled the 200 trajectory ensembles with replacement, 
  averaging the mean MSD curves calculated from each ensemble and then calculating its 
  slope to create a distribution of $D$ values. We report the mean and standard
  error of the distribution of $D$.

@@

  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{qualitative_unclustered_MET2.pdf}
  \caption{We used the parameters of the HDP-AR-HMM, whose state segments are 
  visualized in Figure~\ref{fig:rz_unclustered}, in order to construct an 
  ensemble of characteristic trajectories. The stochastic trajectories 
  generated by the parameters of the HDP-AR-HMM (blue) show 
  %MRS: not sure we can say quite the same 
  %the same 
  similar hopping
  and trapping behavior exhibited by MD (black).
%  We can use realizations of the HDP-AR-HMM to generate trajectories which 
%  bear qualitative resemblance to the MD trajectories. 
  }\label{fig:qualitative_unclustered}
  \end{figure}

  In Figure~\ref{fig:unclustered_msds}, we show that in most cases we can 
  reproduce the magnitude of the MD MSDs on long timescales within uncertainty
  using realizations of the HDP-AR-HMM. We generated the MSD curves using Method
  1 described in Section~\ref{method:realizations} of the methods. Our ability
  to reproduce the MSDs gives us confidence in their long term predictions.
  %MRS: not match under any circumstance, or without clustering (?  Not clear. 
  %BJC5: not exactly sure what you're asking here. This section has no clustering 
  We underestimate the MSD of urea because of a small number of MD trajectories
  with uncharacteristically high MSDs. Despite our best efforts, we could not 
  adequately parameterize an HDP-AR-HMM which reproduced the uncharacteristic
  behavior (see SI).  %BJC5: TODO. There are a couple trajectories I want to show have very high MSDs because all the hops are in one direction along the pore. The curves look superdiffusive. I think on longer timescales, this wouldn't happen. I'll also show the MD MSD curves without the 'outliers' in the SI.
  On long time scales, the model's MSD appear to have a slope similar to the MD 
  MSD.

  %BJC5: need to update legend to say HDP-AR-HMM
  \begin{figure}
  \centering
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET.pdf}
  \caption{methanol}\label{fig:unclustered_msd_MET}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_GCL.pdf}
  \caption{ethylene glycol}\label{fig:unclustered_msd_GCL}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_URE.pdf}
  \caption{urea}\label{fig:unclustered_msd_URE}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_ACH.pdf}
  \caption{acetic acid}\label{fig:unclustered_msd_ACH}
  \end{subfigure}
  \caption{For methanol, ethylene glycol and acetic acid, the MD MSDs (black) and the MSDs
  calculated from realizations of the HDP-AR-HMM match within uncertainty. The MD MSD of
  urea is under-predicted by the HDP-AR-HMM.}\label{fig:unclustered_msds}
  \end{figure}

>>>>

  Appropriate time series modeling of complex diffusion in soft matter systems on the
  microsecond time scale can provide a path towards inferring transport mechanisms and
  predicting bulk properties characteristic of much longer time scales. In this work 
  we apply the sticky hierarchical Dirichlet process autoregressive hidden Markov model
  (HPD-AR-HMM) to solute center-of-mass trajectories generated from long molecular dynamics (MD)
  simulations in a cross-linked inverted hexagonal phase lyotropic liquid crystal (LLC)
  membrane in order to automatically detect a variety of solute dynamical modes.

  We can better understand the mechanisms controlling these dynamical modes by grouping 
  the states identified by the HDP-AR-HMM into clusters based on multiple metrics
  aimed at distinguishing solute behavior based on their fluctuations, dwell times
  in each state and position within the inhomogeneous membrane structure. We analyze
  prevalent clusters in order to relate their parameters to physical interactions between 
  solutes and the membrane.

@@

  % BJC4: even if we could calculate diffusion constants, wouldn't there still be value 
  % in understanding chemical detail?
  % BJC4: It might not be a good idea to blame our need for this modeling on non-linear MSDs, 
  % especially since our MSDs appear to be at or nearing a linear regime.
  % BJC4: seems like we should be addressing the problem as short time scale subdiffusive 
  % dynamics that dictates measurable long time scale diffusion. Although curves become linear,
  % the slopes should be a product of the short time scale motion.
  %MRS: OK, did a pass below, reorganizing it a bit.  See what you think. I think it's OK to talk
  %about subdiffusion because it's general problem, we aren't making claims about this one. 
  %BJC6: looks good!

  The ability to efficiently design novel membrane materials for selective separations 
  would be greatly enhanced by easily interpretable methods for connecting microscopic
  fluctuations with macroscopic observables. If one can use molecular dynamics (MD)
  simulations to explain important experimental metrics such as solute flux and 
  selectivity in terms of detailed chemically dependent solute motion, then we can 
  more easily bridge the disconnect between theory and application. In many simple 
  systems, short-time scale dynamics and large scale properties can be connected by
  the simple relationship between the mean square displacement (MSD) of the solutes
  and the macroscopic diffusion constant. However, for systems which exhibit complex 
  short timescale subdiffusive dynamics, one cannot reliably compute diffusion 
  constants because MSDs are often non-linear on the timescales accessible to molecular
  simulation. Additionally, direct calculations of diffusion constant may obscure the
  molecular mechanisms controlling transport, hindering the ability to understand or 
  modulate these mechanisms in new membrane materials. We are therefore in need of a 
  more flexible approach for modeling solute dynamics which makes less assumptions 
  about the solute's short time scale behavior.

@@

  In this work we attempt uncover similar physical meaning through a data-driven process. 
  We utilize the sticky hierarchical Dirichlet process autoregressive hidden Markov model 
  (HDP-AR-HMM), a nonparameteric Bayesian classification technique that allows the user to 
  automatically detect and infer the parameters of an unknown number of latent vector
  autoregressive (VAR) modes present in solute center-of-mass MD 
  trajectory.~\cite{fox_bayesian_2010} The sticky HDP-AR-HMM greatly extends the 
  basic framework of simpler but more limited finite hidden Markov modeling approaches.

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm, an expectation maximization technique, is a commonly 
  used way to determine the hidden state sequence and the underlying parameters of
  a standard HMM,~\cite{baum_maximization_1970} though it is also possible to achieve the same
  results using a Bayesian framework.~\cite{scott_bayesian_2002,jasra_markov_2005} 
  One can also place priors on the transition probabilities and parameters of the hidden 
  states' emission distributions in order to iteratively infer their values and to
  estimate the hidden state sequence. HMMs have been used in a broad range of applications
  including speech recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}.
  In molecular simulations, HMMs have been used to characterize molecular kinetics
  in order to determine the equilibrium distribution of various observables.~\cite{thayer_hidden_2002,singhal_using_2004,noe_probability_2008,noe_projected_2013}

@@

  To our knowledge, applying the HDP-AR-HMM in order to gain kinetic and
  mechanistic insight from MD simulation trajectories has not been done
  before. Typical applications of the HDP-AR-HMM center around determining the 
  number of states, the state sequence and occasional analysis of the emission 
  distributions of a time series. The addition of temporal correlation to 
  the inference procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  In our approach, we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, and we cluster and analyze the temporally-correlated 
  parameters of the model in order to learn physical mechanisms and describe
  solute kinetics. We can use the model parameters determined either before 
  or after clustering to generate new time series trajectories which enable 
  projections of the MD-generated time series to arbitrary length time scales.

  We aim to address two primary scientific questions with this work:
  \begin{enumerate}
  \item First, we want to know whether it is possible to use the HDP-AR-HMM to 
  help us predict macroscopic transport properties. We use the VAR parameters
  and state transition probability matrix of the HDP-AR-HMM in order to 
  generate stochastic trajectory realizations with behavior similar to that
  exhibited by MD. One can project these realizations onto much longer 
  time scales with computational ease extending into time scales where 
  regular diffusive behavior again occurs.
  \item Second, we want to know if the HDP-AR-HMM algorithm can help
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior.
  We use the parameters of the states identified by the HDP-AR-HMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.
  \end{enumerate}

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10\% water by weight. Approximately one third of the water molecules 
  occupy the tail region with the rest near the pore center.~\cite{coscia_chemically_2019}

  We chose to study a subset of 4 of the fastest moving solutes from our previous
  work: methanol, acetic acid, urea and ethylene glycol. In addition to exploring 
  membrane structural space the most, these solutes have a relatively diverse set
  of chemical functionality. For each solute we created a separate system and to 
  each system we added 6 solutes per pore for a total of 24 solutes. This number 
  of solutes per pore provides a sufficient amount of data from which to generate 
  statistics on the time scales which we simulate and prevents solute clustering
  facilitating our study of pure solute-membrane interactions. Further details on
  the setup and equilibration of these systems are detailed in our previous work.~\cite{coscia_chemically_2019}

@@

  We use the sticky HDP-AR-HMM in order to assign the time-correlated positional 
  fluctuations of solute center-of-mass MD trajectories to discrete dynamical modes while 
  simultaneously inferring the VAR parameters of those modes. We use the sticky model
  in order to emphasize self-transitions since we expect solutes to stay trapped in
  states for appreciable amounts of time.

@@

  \textit{Parameterizing the Hidden States}: We describe the dynamics of each 
  state visited by solutes in our MD simulations using a first order vector 
  autoregressive (VAR(1)) model. In general, a VAR($r$) process is characterized by a
  vector of observations in a time series that are linearly dependent on $r$ previous
  values of the time series vector:
  \begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  \label{eqn:var}
  \end{equation}
  Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  process is further characterized by a shift in the mean of each dimension by the
  vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  covariance, $\Sigma$. We limited our analysis to an autoregressive order of $r=1$.
  This means that we only parameterize $A_1$. To simplify notation, we will call it $A$. 
  We used a conjugate matrix-normal inverse-Wishart (MNIW) prior on parameters 
  $A$ and $\Sigma$ and a Gaussian prior on $\mathbf{c}$ in order to efficiently infer their 
  values.~\cite{fox_nonparametric_2009}

@@

  We ran 2000 iterations of the HDP-AR-HMM procedure on each trajectory which
  was sufficient in order to arrive at converged state sequences and VAR parameter
  estimates. By our criteria, the procedure generally converged within 500--1000 iterations. We 
  evaluated convergence based on when the VAR parameter estimates plateaued. 
  We detected plateaus using the module \texttt{pymbar.timeseries.detect\_equilibration}~\cite{chodera_simple_2016} 
  on the time series of the individual parameters (Section~\ref{S-section:convergence}
  of the Supporting Information). Convergence of the VAR parameters implies that
  the state sequence is relatively stable since the inferred parameters are 
  sensitive to the inputs to the MNIW conjugate priors. The precise boundaries of 
  the state segments can fluctuate; therefore, in states with a low number of 
  emissions, the variance in $A$ and $\Sigma$ parameters can be high. In these 
  cases, plateauing of the mean in each dimension tends to be a more reliable 
  indicator of convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting
  Information).

@@

  There are alternative ways to cluster the data dependent on one's goals. Our goal is
  to derive the simplest set of parameters which adequately describe solute motion. 
  However, one may want a more detailed understand of specific dimensions of the 
  clustering dataset. In this case one can cluster on one or more dimensions independently
  and then enumerate the possible clusters. This would ensure very little overlap of
  clusters but may result in a large number of them. Once one has achieved a better 
  understanding of solute-membrane interactions, perhaps by using the HDP-AR-HMM, it may be
  beneficial to incorporate other physically motivated variables into the clustering
  procedure.

@@

  \textit{Obtaining Parameters of Clustered States}: We remapped and fixed the
  state sequence based on the cluster assignments and 
  parameterized the clusters using the inference component of the HDP-AR-HMM algorithm.
  First, we modified the ($x$, $y$, $z$) solute trajectories so that they had a 
  mean of zero, leaving only the fluctuations. We took advantage of the
  rotational invariance of the system by rotating the coordinates about the pore 
  axis such that the mean ($x$, $y$) vector of each separate segment was aligned with the 
  $x$ axis, forcing the $y$ component to zero. Next, we subtracted the means from 
  each separate segment in the $x$ and $z$ dimensions. We used
  the inference component of the HDP-AR-HMM algorithm on this modified trajectory to 
  infer the clustered state parameters by fixing the clustered state sequence. See
  Section~\ref{S-section:ihmm_procedure} for figures illustrating this process.

@@

  \subsection{Predicting Mean Squared Displacement}\label{method:realizations}

  \textit{Generating Stochastic HDP-AR-HMM Trajectory Realizations}: We used two 
  different methods to stochastically generate trajectories because generating
  trajectories from clustered parameter sets can give results that are qualitatively 
  and quantitatively different from those generated by unclustered parameter sets.
  In our discussion we evaluate both methods:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
    \item For each independent MD trajectory, we generated stochastic trajectory
    realizations using the final parameters of the associated HDP-AR-HMM. This 
    results in 24 separate ensembles of trajectories for each solute studied. 
    \item Using the clustered parameter sets, we generated a single ensemble
	of trajectories. Realizations are representative of the combined behavior
	of all solute trajectories.
  \end{enumerate}

@@

  \subsection{Predicting Selectivity}\label{method:selectivity}

@@

  Solute flux, as written, can be expressed in terms of the product of solute 
  diffusivity, $D_i$, within the membrane and a sorption coefficient, $K_i$, 
  that quantifies the ability of the solute to move between the bulk solution
  and and the membrane:
  \begin{equation}
    J_i = \frac{D_i K_i}{l}\Delta C
  \end{equation}
  where $l$ is the membrane thickness. For our purposes, $K_i$ and $l$ are 
  irrelevant since we do not study pore entry effects or flux across finite
  length pores. Therefore, in order to calculate selectivity, with respect
  to quantities we can study with our MD simulations, it is only necessary that
  we compare the ratio of diffusion constants:
  \begin{equation}
    S_{ij} = \frac{D_i}{D_j}
    \label{eqn:selectivity_diffusivity}
  \end{equation}
  It can be shown on time scales where motion is uncorrelated, that the diffusion
  constant is proportional to the slope of the linear region of the MSD 
  curve:~\cite{einstein_investigations_1956}
  \begin{equation}
    \overline{z^2(\tau)} = 2 n D t
    \label{eqn:msd_D}
  \end{equation}
  where $n$ is the number of spatial dimensions (1 in our case). In general, the 
  MSD of the realizations of our models enters a linear regime near a 5 ns time lag
  (see Figure~\ref{S-fig:MET_nclusters} of the Supporting Information for example).
  Since MSD sampling decreases with increasing time lag, longer trajectory 
  realizations help to extend the linear region.
  %BJC: need to confirm these details
  For all solutes, we generated 200 ensembles of 24 5 $\mu s$ trajectories and 
  estimated $D$ from the slope fit between 10 ns and 1000 ns. We used weighted 
  least squares to fit the slope with weights determined by the number of time 
  lags used to compute the average ($T - \tau$ in equation~\ref{eqn:tamsd}). 
  We bootstrapped error bars for the final reported selectivities. For each 
  bootstrap trial, we sampled the 200 trajectory ensembles with replacement, 
  averaging the mean MSD curves calculated from each ensemble and then calculating its 
  slope to create a distribution of $D$ values. We report the mean and standard
  error of the distribution of $D$.

@@

  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{qualitative_unclustered_MET2.pdf}
  \caption{We used the parameters of the HDP-AR-HMM, whose state segments are 
  visualized in Figure~\ref{fig:rz_unclustered}, in order to construct an 
  ensemble of characteristic trajectories. The stochastic trajectories 
  generated by the parameters of the HDP-AR-HMM (blue) show similar hopping
  and trapping behavior exhibited by MD (black).
  }\label{fig:qualitative_unclustered}
  \end{figure}

  In Figure~\ref{fig:unclustered_msds}, we show that in most cases we can 
  reproduce the magnitude of the MD MSDs on long timescales within uncertainty
  using realizations of the HDP-AR-HMM. We generated the MSD curves using Method
  1 described in Section~\ref{method:realizations} of the methods. Our ability
  to reproduce the MSDs with stochastic models learned from the trajectories gives us confidence in their long term predictions.
  We underestimate the MSD of urea because of a small number of MD trajectories
  with uncharacteristically high MSDs. Despite our best efforts, we could not 
  adequately parameterize an HDP-AR-HMM which reproduced the uncharacteristic
  behavior (see Section~\ref{S-section:urea_underestimate} of the Supporting Information).
  On long time scales, the model's MSD appear to have a slope similar to the MD 
  MSD.

  %BJC6: added. Should I add a figure?
  A potential shortcoming of the HDP-AR-HMM is its inability to reproduce the 
  short time scale MSD curves. The MSDs predicted by the model tend to enter
  a linear regime much more quickly than the MD MSDs (see 
  Figure~\ref{S-fig:unclustered_msd_shortlag} of the Supporting Information
  for example). This is not a major concern for predicting macroscopic properties
  since we are primarily interested in reproducing the MSDs on much longer 
  timescales.
  %BJC6: could hypothesize why using previous work as support

  \begin{figure}
  \centering
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_MET.pdf}
  \caption{methanol}\label{fig:unclustered_msd_MET}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_GCL.pdf}
  \caption{ethylene glycol}\label{fig:unclustered_msd_GCL}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_URE.pdf}
  \caption{urea}\label{fig:unclustered_msd_URE}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{unclustered_msd_ACH.pdf}
  \caption{acetic acid}\label{fig:unclustered_msd_ACH}
  \end{subfigure}
  \caption{For methanol, ethylene glycol and acetic acid, the MD MSDs (black) and the MSDs
  calculated from realizations of the HDP-AR-HMM match within uncertainty. The MD MSD of
  urea is under-predicted by the HDP-AR-HMM.}\label{fig:unclustered_msds}
  \end{figure}
